Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2019-03-11 accessPEP.Rmd
cleaning up getting data and running models

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## Libraries
library(data.table)
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
library(stringr)
library(ISOweek)
library(raster)
library(sp)
library(MASS)
library(reshape2)
source("R/utils.R")

select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
                      results = "hide",
                      include = TRUE)
```

## INTRODUCTION

Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis. However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*). Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.

Here--discuss geographic variation in access to vaccination and care--how this shapes mortality for other diseases.

The majority of rabies burden studies use these data to estimate burden from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space. While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites. The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level. Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.

@@

## Getting commune covars
## Need: weighted ttimes, pop, ctar_in, exclude
comm_catchments %>%
  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) %>%
  left_join(select(mada_district@data, mdg_dis_co, district, ctar_in_dist, dist_catch = CTAR, exclude,
                   color, fill, ctar_dist)) -> comm_data
comm_data <- filter(comm_data, ttimes_weighted != "Inf")

# ## Data to include
# nrow(dist_catchments) - sum(dist_catchments$exclude)
# nrow(comm_data) - sum(comm_data$exclude)

#### Moramanga data! ###
## ctar for # bites and also checking days to rule out contacts
col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))

## Set-up time series
tfunc <- function (date, format_date = "%Y-%m-%d", start = "01-09-2016", format_start = "%d-%m-%Y",
                   year1 = 2016, tstep = "month", period = FALSE) {
  get.consec (date, format_date, start, format_start,
              year1, tstep, period) - 8 ## starts in Sep
}

## all dates
ctar %>%
  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
ctar$month <- tfunc(ctar$date.reported)
## subset to study period (Sep 2016 - to date)
## checks for cases where month is NA!
ctar <- subset(ctar, month > 0 & month < 28)

# ctar district and commune
ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
ctar$commune <- trimws(ctar$commune, which = "right")
ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
ctar$district <- gsub(" \\(District\\)", "", ctar$district)

## change all true/false to 0/1
ctar[ctar == FALSE] <- 0

## Bite data in Moramanga
n_months <- max(ctar$month, na.rm = TRUE)

ctar %>%
  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
            by = c("commune" = "commune", "district" = "mtch_ds")) %>%
  filter(consult.type == "Animal bite/exposure") %>%
  group_by(mdg_cm_) %>%
  summarize(bites = round(n()/n_months*12)) %>%
  left_join(comm_data) %>%
  filter(!is.na(ttimes_weighted), 
         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites
```

@@

```{r network fig}
## Network fig
## Revise: points = CTAR points with size proportional to number of bites reported, lines to district proportional to the number of bites from that district, outline = by district, communes = polygons shaded by catchments

## get exposure matrix = first set of points
## district exposures
ctar_data %>% 
  drop_na(id_ctar, district) %>%
  group_by(district) %>% # group by district
  summarize(n = n()) -> exps_dist # get count of exposures
## ctar exposures
ctar_data %>% 
  drop_na(id_ctar) %>%
  group_by(id_ctar) %>% # group by ctar
  summarize(n = n()) -> exps_ctar
## ctar by district exposure matrix
ctar_data %>% 
  drop_na(id_ctar, district) %>%
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat# get count of exposures

## To plot district bites
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(LONGITUDE, dist_long), 
        dist_lat = coalesce(LATITUDE, dist_lat)) -> dist_points

exps_dist %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

## To plot ctar points (only those not excluded)
exps_ctar %>%
 left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE, 
                  exclude, color, fill)) %>%
 filter(exclude == 0) -> ctar_points

## To plot lines between ctar + districts
exp_mat %>% 
  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.25
```

```{r figure 1A}
jpeg("figs/network.jpeg")
plot(mada_district, col = mada_district$fill, border = "grey") 
segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, 
         y1 = dist_lines$LATITUDE,
         lwd = dist_lines$size*2, col = dist_lines$color)
points(ctar_points$LONGITUDE, ctar_points$LATITUDE, col = ctar_points$color, pch = 1,
       cex = log(ctar_points$n + 0.1)*0.25)
points(dist_points$dist_long, dist_points$dist_lat, cex = dist_points$size, pch = 20,
       col = dist_points$color)
legend("topright", c("10", "50", "100", "200", "400", "800", "1600"), pch = 1, xjust = 1, yjust = 1,
       pt.cex = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, 
       lwd = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, bty = "n", y.intersp = 0.75,
       title = "Number of \n reported bites", xpd = TRUE)
legend("topleft", "A", bty = "n", inset = c(0.2, 0), text.font = 2)
dev.off()
```

```{r network output, echo=FALSE, fig.cap="Figure 1A", results = "show", out.width = '100%'}
knitr::include_graphics("figs/network.jpeg")
```

```{r figure 1b, fig.cap = "Figure 1B"}
exp_mat %>% 
  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
            by = c("district" = "mdg_dis_co")) %>%
  group_by(district) %>% # group by ctar
  mutate(prop = n/sum(n),
         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
  filter(catch_match == 1) -> catch_dist
hist(catch_dist$prop, main = "", 
     xlab = "Proportion of patients \n reporting to closest ARMC",
     ylab = "Number of districts", col = "grey50", border = "white", xlim = c(0, 1))

exp_mat %>% 
  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
            by = c("district" = "mdg_dis_co")) %>%
  group_by(ctar_obs) %>% # group by ctar
  mutate(prop = n/sum(n),
         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
  filter(catch_match == 1) %>%
  summarize(prop = sum(prop)) -> catch_ctar
```

```{r figure 1C, fig.cap = "Figure 1C"}
hist(catch_ctar$prop, main = "", 
     xlab = "Proportion of patients reporting from within catchment",
     ylab = "Number of ARMC", col = "grey50", border = "white", xlim = c(0, 1))
```

@@

```{r clinic level reporting, fig.cap = "Figure 2A"}
######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  drop_na(id_ctar) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts

patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")

## getting reporting
ts <- as_tibble(seq(start_date, end_date, by = "day"))

patient_ts %>%
  right_join(ts, by = c("date_de_consultation" = "value")) %>%
  select(id_ctar, date_de_consultation, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

get.days <- function(dmat = dose_mat, threshold = 10) {
  date_mat <- matrix(NA, nrow(dmat), ncol(dmat))
  for (j in 1:ncol(date_mat)){
    rle(dmat[ , j]) %>%
      unclass() %>%
      as.data.frame() %>%
      mutate(end = cumsum(lengths),
             start = c(1, lag(end)[-1] + 1)) %>%
      filter(values == 0, lengths >= threshold) -> rles
  if(nrow(rles) > 0){
    for (i in 1:nrow(rles)){
        date_mat[rles$start[i]:rles$end[i], j] <- 0
      }
    }
  }

  date_mat <- replace(date_mat, is.na(date_mat), 1)
  date_mat %>% 
    as_tibble() %>%
    group_by(date = year(doses_wide$date_de_consultation)) %>%
    summarise_all(funs(sum(.)/n())) -> clinic_reporting
  colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]

  clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
  clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
  clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
  clinic_reporting$threshold <- threshold
  return(list(clinic_reporting, date_mat))
}

date_mat <- as.data.frame(get.days(threshold = 15)[[2]])
date_mat$date_de_consultation <- doses_wide$date_de_consultation
date_mat <-melt(date_mat, id = "date_de_consultation")
levels(date_mat$variable) <- names(doses_wide)[2:23]
names(date_mat) <- c("date_de_consultation", "id_ctar", "exclude")

patient_ts %>%
  ungroup() %>%
  mutate(id_ctar = as.factor(id_ctar)) %>%
  right_join(date_mat) %>%
  mutate(n = replace_na(n, 0), 
         n = ifelse(exclude == 1, n, NA)) -> doses_long
doses_long$ctar <- ctar_metadata$CTAR[match(doses_long$id_ctar, ctar_metadata$id_ctar)]

ggplot(data = doses_long, aes(date_de_consultation, ctar)) + 
   geom_tile(aes(fill = n)) +
   scale_fill_gradientn(colours = c("white", "purple", "black"), values = c(0, 0.1, 1),
                        na.value = alpha("lightgrey", 1)) +
   xlim(start_date, end_date)

clinic_reporting <- get.days(threshold = 15)[[1]]

```{r figure 1B, fig.cap = "Figure 1B"}
clinic_reporting %>%
  mutate(prop15 = prop, prop5 = get.days(threshold = 5)[[1]]$prop,
         prop30 = get.days(threshold = 30)[[1]]$prop) %>%
  ggplot(., aes(x = reorder(ctar, prop), color = ctar)) +
  geom_hline(yintercept = 0.25, color = "darkgrey") +
  geom_boxplot(aes(ymin = prop30, lower = prop15, middle = prop15, upper = prop15, ymax = prop5),
               stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  xlab("") +
  ylab("Proportion of days included") +
  coord_flip() +
  labs(tag = "B")

```{r figure 3B, fig.cap= "Figure 3B"}
patient_ts %>%
  group_by(ctar) %>%
  mutate(exclude = ifelse(n > mean(n) + 3*sd(n), 1, 0)) -> patient_ts

ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
  scale_color_discrete(name = "Exclude", labels = c("Yes", "No")) +
  ylab("Number of patients per day") +
  geom_jitter(alpha = 0.5, width = 0.25) +
  coord_flip()

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$exclude_bycomment <- 0
ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
patient_ts %>%
  ungroup() %>%
  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
  right_join(ctar_data) -> ctar_check

```{r trying to exclude contacts, fig.cap="Figure 3A"}
## Testing with Moramanga data 
sdev <- seq(2, 10, by = 1)
check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
for (i in 1:length(sdev)){
  ctar %>%
    group_by(date.reported) %>%
    summarize(n = n()) %>%
    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
  check <- rbind(check, c(sdev[i], checksd[1:2]))
}

plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "n", ylim = c(0, 1), 
     ylab = "Proportion excluded", xlab = "Number of standard deviations")
points(check$sd, check$prop_bites, col = "red", pch = 20)
legend("topright", c("Contacts", "Bites"), col = c("Blue", "Red"), pch = 20, bty = "n")
```

@@

```{r district level exposures, fig.cap = "Figure 4"}
## districts that are non-reporter and data not available yet CTAR
mada_district@data %>%
select(district = mdg_dis_co, dist_name = district, pop = pop2015adj, ttimes_weighted_dist = ttimes_weighted, 
       dist_catch = CTAR, exclude, color) -> covars

ctar_check %>% 
  filter(exclude_bydate == 0, exclude_bycomment == 0) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
  summarize(n = n()) %>% # get count of exposures
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) %>% ## filter out excluded catchments and ones with < 25% reporting
  group_by(district, year) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  left_join(covars) -> exps

col_vals <- dist_catchments$color
names(col_vals) <- dist_catchments$mdg_dis_co

ggplot(exps, aes(x = reorder(dist_name, ttimes_weighted_dist), y = bites/pop*1e5, color = district)) +
  geom_boxplot() +
  theme(legend.position = "none") +
  scale_color_manual(values = col_vals) + 
  ylab("Annual bites per 100k") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

exps %>%
  group_by(district) %>%
  summarise(bites = mean(bites), n = n()) %>%
  left_join(covars) -> exps_dist
```

@@

### Models of bites as a function of travel times
```{r likelihood of observing moramanga data @ commune level, fig.cap = "Figure 5A"}
library(boot)

get.likelihood.bites <- function(bites, covar, pop, params) {
  beta <- params[1]
  intercept <- params[2]
  exp_bites <- inv.logit(beta*covar + intercept)*pop
  return(-sum(dpois(round(bites), lambda = exp_bites, log = TRUE)))
}

get.likelihood.mora <- function(commune, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  commune %>%
      mutate(bites_mu = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) -> sum_bites
  return(-sum(dpois(round(sum_bites$), lambda = sum_bites$bites_mu, log = TRUE)))
}

get.likelihood.mada <- function(commune, district, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  commune %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = inv.logit(B_ttimes*ttimes_weighted + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
}
get.likelihood.dist <- function(district, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  district %>%
      filter(exclude == 0) %>%
      mutate(exp_bites = inv.logit(B_ttimes*ttimes_weighted_dist + B_0)*pop) -> district
  return(-sum(dpois(round(district$exp_bites), lambda = round(district$bites), log = TRUE)))
}

### Moramanga model ###
mora_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mora, commune = mora_bites, hessian = TRUE)
B_ttimes_mora <- mora_pars$par[1]
B_0_mora <- mora_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(mora_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(mora_pars$hessian)

ntest = 1000; testpar = mora_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, mora_pars$par, varcov)
}

B_ttimes_mora_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_mora_CI <- quantile(sim.par[,2],c(0.025,0.975))

### Madagascar model ###
mada_pars <- optim(par = c(1e-6, 0.1), get.likelihood.mada, district = exps_dist,
             commune = comm_data, hessian = TRUE)
B_ttimes_mada <- mada_pars$par[1]
B_0_mada <- mada_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(mada_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(mada_pars$hessian)

ntest = 1000; testpar = mada_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, mada_pars$par, varcov)
}

B_ttimes_mada_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_mada_CI <- quantile(sim.par[,2],c(0.025,0.975))

### District parameters
dist_pars <- optim(par = c(1e-6, 0.1), get.likelihood.dist, district = exps_dist,
              hessian = TRUE)
B_ttimes_dist <- dist_pars$par[1]
B_0_dist <- dist_pars$par[2]

@@

## Mada
par_df <- as.data.frame(list(model = c("Mada communes", "Moramanga communes", "Mada districts"),
                     B_ttimes = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist),
                     B_0 = c(B_0_mada, B_0_mora, B_0_dist),
                     B_0_upper = c(B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                     B_0_lower = c(B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]),
                     B_ttimes_upper = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_ttimes_dist_CI[2]),
                     B_ttimes_lower = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  B_ttimes_dist_CI[1])))

>>>>

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## Clear memory
rm(list = ls())
## Libraries
library(data.table)
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
library(stringr)
library(ISOweek)
library(raster)
library(sp)
library(MASS)
library(reshape2)
library(geosphere)
library(foreach)
library(bbmle)
library(boot)
library(geosphere)

## Files
source("R/utils.R")
source("R/models.R")
source("R/data_functions.R")
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, 
                      results = "hide",
                      include = TRUE)
```

## INTRODUCTION
Inequities in access to care are a major driver of disease burden globally. Often, the populations at greatest risk of a given disease are the most underserved. Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful considerations of how best to optimize the allocation of limited resources.

Canine mediated rabies is estimated to cause approximately 60,000 human deaths anually. These deaths can be prevented through prompt administration of post-exposure prophylaxis. However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*). GAVI bit here...potential to greatly reduce disease burden.

Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities. The majority of rabies burden studies thus far have integrated these data on reported bites into a probability decision tree framework to estimate rabies burden, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space. While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites.

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.

@@

## Getting commune covars
## Need: weighted ttimes, pop, ctar_in, exclude
ctar_coords <- cbind(ctar_metadata$LONGITUDE, ctar_metadata$LATITUDE)
comm_catchments %>%
  left_join(select(mada_communes@data, pop = MDG__201, mdg_cm_, district = mtch_ds)) %>%
  left_join(select(mada_district@data, mdg_dis_co, district, ctar_in_dist, dist_catch = CTAR, exclude,
                   color, fill, ctar_dist)) -> comm_data
mada_communes@data %>%
  mutate(long = coordinates(mada_communes)[, 1], 
         lat = coordinates(mada_communes)[, 2]) %>%
  select(commune, mdg_cm_, mtch_ds, long, lat) %>%
  right_join(comm_data) %>%
  mutate(distance = apply(distm(select(., long, lat), ctar_coords), 1, min)/1000) -> comm_data
comm_data <- filter(comm_data, ttimes_weighted != "Inf")

## Getting district data
mada_district@data %>% 
  select(mdg_dis_co, district, pop = pop2015un, ttimes_weighted, ctar_in_dist, 
         dist_catch = ctar_dist, exclude, color, fill) -> dist_data

exps_dist <- get.district.data(ctar_data = ctar_data, ctar_metadata = ctar_metadata, 
                               mada_district = mada_district, catchments, 
                               consec_days = 15, 
                               contact_cutoff = 3, reporting_thresh = 0.25, 
                               start = ymd("2014-01-01"), end = ymd("2017-12-31"))
exps_mora <- get.morabites(mada_communes = mada_communes)

@@

@@

@@

@@

### Models of bites as a function of travel times

```{r models}
model.bites <- function(bites, names_bites, covar, pop, names_covar, sum = TRUE, 
                        pop_predict = "addPop", covar_name = "ttimes_weighted",
                        beta, beta_pop, intercept, trans, run = "optim",...)

## Exclude for comm data!
comm_covars <- filter(comm_data, exclude == 0)

pop_predict <- c("addPop", "onlyPop", "none")
pop_vars <- list(pop_dist = exps_dist$pop, pop_commune = comm_covars$pop)
covars <- list(distance_district = exps_dist$distance, distance_commune = comm_covars$distance, 
               ttimes_district = exps_dist$ttimes_weighted_dist/60, ttimes_commune = comm_covars$ttimes_weighted/60)
sum_it <- c(FALSE, TRUE, FALSE, TRUE)
run_type <- "optim"

mods_mada <- 
  foreach(i = 1:length(pop_predict), .combine = c) %:%
  foreach(j = 1:length(sum_it), .combine = c) %:%
  foreach(k = 1:length(covars), .combine = c) %do% {
  i = 1; j = 1; k = 1;

  if(sum_it[j] == TRUE) {
    names_covar <- comm_covars$mdg_dis_co
  } else {
    names_covar = ""
  }
  data_est <- list(bites = exps_dist$bites, run = run_type, covar = covars[[k]],
                   names_bites = exps_dist$district,
                   covar_name = names(covars)[k], names_covar = names_covar,
                   trans = 1e5)
  start_params <- list(beta = 1e-5, intercept = 0.1)

  if(pop_predict[i] == "addPop") {
    ## need beta_pop
    start_params <- c(start_params, beta_pop = 1e-5)
  }

  if(grepl("commune", names(covars)[k], fixed = TRUE)) {
    data_est <- c(data_est, list(pop = pop_vars$pop_commune))
  } else {
    data_est <- c(data_est, list(pop = pop_vars$pop_dist))
  }

  check <- mle2(model.bites, start = start_params, data = data_est)

  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
  check
}

mods_moramanga
mods_pop_mada_moramanga

## all mods = combine

prof_CI <- lapply(mods, confint)

mada_mod_predicts <- as.data.frame(list(beta = mapply(function(x) coef(x)["beta"], mods), 
     intercept = mapply(function(x) coef(x)["intercept"], mods), 
     beta_upper = mapply(function(x) x["beta", 2], prof_CI),
     beta_lower = mapply(function(x) x["beta", 1], prof_CI),
     intercept_upper = mapply(function(x) x["intercept", 2], prof_CI),
     intercept_lower = mapply(function(x) x["intercept", 1], prof_CI),
     likelihood = mapply(function(x) logLik(x), mods),
     model = mapply(function(x) attributes(x)$name, mods)))

mada_mod_results <- as.data.frame(list(params = c("beta", "intercept"), 
                   estimates = t(mapply(function(x) coef(x), mods)), 
                   upper = t(mapply(function(x) x[, 2], prof_CI)),
                   lower = t(mapply(function(x) x[, 1], prof_CI)), 
                   data = "Mada"))

## Predictions
library(foreach)

mods <- 
  foreach(l = 1:nrow(mada_), .combine = c) %do% {
      if(pop_predict[k] == TRUE) covar_name = "pop" else covar_name = names(covar_mada)[l]
      if (scale[i] == "Mada") {
        if(sum[j] == TRUE) {
          sum_name <- "summed"
          check <- predict.bites(bites = exps_dist$bites, 
                           names_bites = exps_dist$district, 
                           pop = exps_dist$pop,
                           pop_sub = comm_data$pop, 
                           names_pop = comm_data$mdg_dis_co, 
                           names_covar = comm_data$mdg_dis_co, trans = 1e5,
                           covar_sub = covar_mada_sub[[l]],
                           sum = sum[j], pop_predict = pop_predict[k])
        } else {
          sum_name <- ""
          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
               data = c(data_mada, list(covar = covar_mada[[l]]),
                           sum = sum[j], pop_predict = pop_predict[k]))
        }
      }   
      if(scale[i] == "Mora") {
        if(sum[j] == TRUE) {
          check <- NULL
        } else {
          sum_name <- ""
          check <- mle2(get.likelihood.bites, start = list(beta = 1e-5, intercept = 0.1),
             data = c(data_mora, list(covar = covar_mora[[l]]),
                           sum = sum[j], pop_predict = pop_predict[k]))
        }
      }
  attributes(check)$name <- paste(scale[i], sum_name, covar_name)
  check
}

ttimes_plot <- seq(0, 720, by = 30)
pop_plot <- seq(100, 1e6, length.out = length(ttimes_plot))
expect <- foreach(i = 1:nrow(mada_mod_predicts), .combine = rbind) %do% {
  check <- mada_mod_predicts$expect_fun[[i]]
  exp <- check(beta = mada_mod_predicts$beta[i], intercept = mada_mod_predicts$intercept[i],
               covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, 
               trans = 1e5)
  exp_upper <- check(beta = mada_mod_predicts$beta_upper[i], 
                     intercept = mada_mod_predicts$intercept_upper[i],
                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
  exp_lower <- check(beta = mada_mod_predicts$beta_lower[i], 
                     intercept = mada_mod_predicts$intercept_lower[i],
                     covar = ttimes_plot/60, pop = 1e5, pop_covar = pop_plot, trans = 1e5)
  as.data.frame(list(ttimes = ttimes_plot, 
                     pop = pop_plot, exp = exp, exp_upper = exp_upper, exp_lower = exp_lower, 
                     model = rownames(mada_mod_predicts)[i], data = mada_mod_predicts$data[i]))
}

ggplot(data = preds, aes(x = observed, y = predicted, color = model)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey")
ggplot(data = filter(expect, !grepl("pop", model, fixed = TRUE)), 
       aes(x = ttimes, y = exp, color = model)) +
  geom_line() +
  geom_ribbon(aes(ymin = exp_lower, ymax = exp_upper, fill = model), alpha = 0.5)

AICctab(mada_ttimes_fit, mada_distance_fit, mada_pop_fit, mada_sumttimes_fit, mada_sumpop_fit,
        mada_sumdist_fit, nobs = nrow(exps_dist))

@@

## Mada
par_df <- as.data.frame(list(model = c("Mada communes", "Moramanga communes", "Mada districts"),
                     B_ttimes = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist),
                     B_0 = c(B_0_mada, B_0_mora, B_0_dist),
                     B_0_upper = c(B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                     B_0_lower = c(B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]),
                     B_ttimes_upper = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_ttimes_dist_CI[2]),
                     B_ttimes_lower = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  B_ttimes_dist_CI[1])))
```
